---
title: "Reproducible Research: Peer Assessment 1"
author: "Dmitry Baranov"
date: "November 2017"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(lattice)
```

## Personal movement activity

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

## Loading and preprocessing the data

Download data from the course web [site](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

```{r loading}
if (!dir.exists("./data"))
    dir.create("./data")

download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "./data/activity.zip")
unzip("./data/activity.zip", exdir = "./data")
activity <- read.table("./data/activity.csv", header= TRUE, sep = ',', stringsAsFactors = FALSE, na.strings = "NA")

head(activity)
```

Transform char dates into Date class
```{r transform dates}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

Histogram of the total number of steps taken each day

```{r hist total number steps per day}
day.activity <- activity %>% group_by(date) %>% summarize(total.steps = sum(steps, na.rm = TRUE))
hist(day.activity$total.steps, breaks = 8, col = "green", xlab = "Number of steps", main = "Total number of steps per day")
```

Mean
```{r mean total number steps per day}
mean(day.activity$total.steps, na.rm = TRUE)
```

Median
```{r median total number steps per day}
median(day.activity$total.steps, na.rm = TRUE)
```

## What is the average daily activity pattern?

Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days
```{r time series total steps}
interval.activity <- activity %>% group_by(interval) %>% summarize(mean.steps = mean(steps, na.rm = TRUE))
with(interval.activity, plot(x = interval, y = mean.steps, type = "l", xlab = "5 min intervals", ylab = "Average number of steps"))
```

Interval with maximum number of steps on average
```{r interval max number steps average}
interval.activity$interval[which.max(interval.activity$mean.steps)]
```

## Imputing missing values

Number of rows with missing values
```{r rows with missing}
nrow(activity[!complete.cases(activity), ])
```

Let's fill in the missing steps values with median values of steps for interval. I suggest that every day in the same interval has similar number of steps in average.
```{r fill in missing values}
interval.median.steps <- activity %>% group_by(interval) %>% summarize(median.steps = median(steps, na.rm = TRUE))
activity.full <- activity %>% inner_join(interval.median.steps, by = "interval") %>% mutate(steps = ifelse(is.na(steps), median.steps, steps)) %>% select(steps, date, interval)
```

Histogram of the total number of steps taken each day with filled missing values

```{r hist total number steps per day fill missing}
day.activity.full <- activity.full %>% group_by(date) %>% summarize(total.steps = sum(steps))
hist(day.activity.full$total.steps, breaks = 8, col = "green", xlab = "Number of steps", main = "Total number of steps per day")
```

Mean with filled missing vales
```{r mean total number steps per day fill missing}
mean(day.activity.full$total.steps)
```

Median with filled missing vales
```{r median total number steps per day fill missing}
median(day.activity.full$total.steps)
```

In results, there is no difference between data sets with and without missing values. Therefore, there is no impact of missing values on the estimates of the total daily number of steps.


## Are there differences in activity patterns between weekdays and weekends?

Indicate date as _weekday_ or _weekend_
```{r indicate week part}
activity.full <- activity.full %>% mutate(weekpart = as.factor(ifelse(weekdays(date, abbreviate = TRUE) %in% c("Sat","Sun"), "weekend", "weekday")))
```
Time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days

```{r time series weekpart avg steps}
weekpart.interval.mean.steps <- activity.full %>% group_by(weekpart, interval) %>% summarize(mean.steps = mean(steps))
xyplot(mean.steps ~ interval | weekpart, data = weekpart.interval.mean.steps, layout = c(1,2), type = "l", ylab = "Number of steps", xlab = "Interval")
```

## My software environment

```{r environment}
sessionInfo()
```
